Assessment of forestry road network projection methods using the “roads” package in R

### 2022-10-14 analysis/paper/paper.Rmd
more edits

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).

When comparing between projection methods we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods. The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results. Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic. The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.

@@

As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance footprint"]`). The cutblocks only method performed worst for all the road network characteristics. For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` for the MST method.

@@

Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data. Often major discrepancies were related to cutblocks that were likely incorrectly labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks. These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections. In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible. Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue. Another potential issue was with roads not being included in the data set that were seen in the satellite imagery. This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets. Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc). This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery. These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn't have and should be treated with caution. It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets. Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context. Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.

Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines). Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc. Because of this, forestry roads make up a much smaller percentage of the overall road network. In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors. The model was attempted in the Fort Nelson TSA however the resulting road network (Figure A-1) did not accurately portray the observed network like it did in the Revelstoke TSA. However, other resource extraction activities could likely be incorporated into this model relatively easily. The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.

@@

## Road network projection methods

Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used three methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST). In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.

LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra's shortest path algorithm (1959). Dijkstra's algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section "weighted graph"). Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

@@

A projection created with the Forest Road Network Plugin for QGIS by Clement (2021) was also created within QGIS and added to the analysis in order to see how it compared with our projection methods. All the default settings such as skidding distance, etc. were left as they were within the QGIS plugin.

@@

A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms. When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. [The edges connecting the nodes follow a "queens" case or allow a maximum of eight possible edges for a given node.]{.underline} Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

@@

```{r figA2, fig.height=6, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of  predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}

>>>>

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

When comparing between projection method variants we found that the density of landing points chosen within cutblocks affected the results, with the high density method underestimating the road density and road presence metrics less than the low density and centroid methods. The sampling method for choosing points in the cutblocks, either random or regular, did not have a large effect on the results. Hardy's QGIS plugin was closest to the observed mean for the road presence and density characteristics but was among the furthest from the observed mean for the distance to the nearest road and road disturbance footprint characteristic. The cutblocks only method showed that not including road development underestimates anthropogenic disturbance.

@@

As expected all the projection method variants preformed worst at projecting the road presence characteristic (F-Measure: `r perf_tbl_rng["Road presence"]`), while the forestry disturbance footprint had the best performance (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`) and road disturbance footprint only slightly worse (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`). The cutblocks only method performed worst for all the road network characteristics (Figure \@ref(fig:fig5)). For the road presence metric higher sampling densities led to somewhat higher F-measure values caused by fewer false negatives and only slightly more false positives. For the forestry and road disturbance footprint metrics, the F-measure was similar across projection method variants, while sensitivity was higher at high sampling densities and precision was higher at low sampling densities. The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

**NOTE David's manuscript rough draft document includes other comments and bullets that might be relevant to rewriting the Intro and Discussion **

## Sarah's thoughts

- The projections work reasonably well for strategic planning purposes, especially if we are not concerned with roads inside cutblocks.

- The differences between algorithms were fairly small and it would be more valuable to focus on the accuracy of input data such as costs and cutblock locations rather than fine tuning the algorithms.

- This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. The results would likely be less accurate if the landings were spread further apart.

## What David had
Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data. Often major discrepancies were related to cutblocks that were likely incorrectly labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks. These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections. In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible. Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue. *Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.* (this is no longer identified in the results section since it wasn't a major discrepancy after fixing the road data source) This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets. Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc). This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery. These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn't have and should be treated with caution. It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets. Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context. Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.

Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines). Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc (\@ref(fig:ftNelson)). Because of this, forestry roads make up a much smaller percentage of the overall road network. In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors. However, other resource extraction activities could likely be incorporated into this model relatively easily. The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.

@@

## Road network projection method details

Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used four methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP), dynamic least cost paths (DLCP) approach and finally, the minimum spanning tree approach (MST). In this paper, we focus on the LCP, DLCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.

The LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra's shortest path algorithm (1959). Dijkstra's algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section "weighted graph"). Witht the LCP approach each STAP is solved independently but for the DLCP approach each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

@@

@@

A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal's or Dijkstra's algorithms. When solving least cost paths using Dijkstra's algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal's algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. The edges connecting the nodes follow an "octagonal" neighbourhood or allow a maximum of eight possible edges for a given node, with diagonal edges having a higher cost. Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design. Therefore, our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

@@

```{r figA2, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Comparison of performance metrics for each projection method variant for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. F measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted."}
